\vspace{2cm} \LARGE \bf Seasonal influenza circulation patterns and projections for 2016-2017

### 2016-09-20 auspice/reports/tex/sep-2016.tex
Update TeX for Sep report.

\date{September 16, 2016}

@@

The late-season 3c3.a epidemic in the USA is puzzling. However, we suspect that this epidemic is due to epidemiologic circumstance rather than the emergence of a selective variant. The strongest evidence for this is that there is not a single clade within 3c3.a that is spreading throughout the USA (Fig.\ \ref{H3N2_3c3a_tree}). Instead, a variety of 3c3.a viruses are spreading, each with different HA1 mutations. The emergence and spread of an adaptive variant would have appeared as a single clade bearing a characteristic epitope mutation. This is not what we see. One small clade, however, carries the S145N and F193S mutations close to the receptor binding site which resulted in antigenic evolution in previous years. The clade is restricted to viruses from US and remains at less than 1\% global frequency. We doubt that these 3c3.a viruses will spread globally. It is conceivable however, that the 2016-2017 USA season could derive from over-summering transmission chains. This would be an unusual event, but not unheard of \cite{bedford2010global}. Still, we believe that 2016-2017 USA season will likely derive from 171K viruses due to their rapid spread (see below).

@@

The dominant 3c2.a viruses have continued to genetically diversify with the emergence of multiple subclades (Fig.\ \ref{H3N2_3c2a_tree}). We observe 3 subclades of decent frequency in 2016 viruses. These are characterized by HA1 mutations 142K/197R, HA1 mutation 197K and HA1 mutation 171K + HA2 mutations 77V/155E. The 171K variant comprises approximately 54\% of 3c2.a viruses collected in 2016.

@@

The 142K/197R clade first emerged around June 2015 and rose to nearly 20\% global frequency in Nov 2015 (Fig.\ \ref{H3N2_mutations}). However, it's declined throughout 2016. We doubt this is a competitive virus. The 197K clade emerged in late 2015 and has slowly grown in frequency since. We estimate that it now comprises $\sim$14\% of H3N2 viruses. However, 197K is at extremely low frequency in Asia with an estimated present-day frequency $\sim$2\%. On the other hand, clade 171K viruses have done remarkably well throughout 2016 and now comprise an estimated 69\% of currently circulating H3N2 viruses. This steady and rapid increase is strongly suggestive of an adaptive origin. Notably, 171K emerged and spread first in Asia, reaching nearly 80\% frequency in April 2016. Higher frequency of 171K in Asia is expected to spread to the rest of the world given historic geographic observations \cite{bedford2015global}.

@@

\textit{Without strong competition from another novel H3N2 virus, we believe that 171K will continue to increase in frequency in the global population and predominate in the 2016-2017 influenza season.} The continued spread of 171K is fully in line with the predictions we made in Feb 2016 \cite{feb2016report}.

@@

@@

\textbf{Clade 3 has continued to predominate the B/Yamagata population. Within clade 3 the 172Q/251Q subclade has risen to predominate. There is limited diversity within 172Q/251Q viruses, with the largest subclade comprised of 211R viruses.}

>>>>

\date{September 20, 2016}

@@

The late-season 3c3.a epidemic in the USA is puzzling. However, we suspect that this epidemic is due to epidemiologic circumstance rather than the emergence of a selective variant. The strongest evidence for this is that there is not a single clade within 3c3.a that is spreading throughout the USA (Fig.\ \ref{H3N2_3c3a_tree}). Instead, a variety of 3c3.a viruses are spreading, each with different HA1 mutations. The emergence and spread of an adaptive variant would have appeared as a single clade bearing a characteristic epitope mutation. This is not what we see. One small clade, however, carries the S145N and F193S mutations close to the receptor binding site which resulted in antigenic evolution in previous years. This is a small clade restricted to viruses from US and remains at less than 1\% global frequency.

@@

We doubt that these 3c3.a viruses will spread globally. It is conceivable however, that the 2016-2017 USA season could derive from over-summering transmission chains. This would be an unusual event, but not unheard of \cite{bedford2010global,bedford2015global}. In one recent season (2008-2009), USA viruses derived primarily from the previous USA season. In other years (at least back to 2000), USA seasons were primarily reseeded from elsewhere (Fig.\ \ref{H3N2_historical_tree_na}). We believe that the 2016-2017 USA season will likely derive from 171K viruses due to their rapid spread (see below).

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/sep-2016/H3N2_historical_tree_na.png}
	\caption{\textbf{Historical phylogeny of H3N2 viruses from North America from Bedford et al.\ (2015) \cite{bedford2015global}.}
	}
	\label{H3N2_historical_tree_na}
\end{figure}

The dominant 3c2.a viruses have continued to genetically diversify with the emergence of multiple subclades (Fig.\ \ref{H3N2_3c2a_tree}). We observe 3 subclades of decent frequency in 2016 viruses. These are characterized by HA1 mutations 142K/197R, HA1 mutation 197K and HA1 mutation 171K + HA2 mutations 77V/155E. The 171K variant comprises approximately 58\% of 3c2.a viruses collected in 2016. Within 171K, the variant 121K has emerged and comprises approximately 23\% of 3c2.a viruses collected in 2016.

@@

The 142K/197R clade first emerged around June 2015 and rose to nearly 20\% global frequency in Nov 2015 (Fig.\ \ref{H3N2_mutations}). However, it's declined throughout 2016. We doubt this is a competitive virus. The 197K clade emerged in late 2015 and has slowly grown in frequency since. We estimate that it now comprises $\sim$14\% of H3N2 viruses. However, 197K is at extremely low frequency in Asia with an estimated present-day frequency $\sim$2\%. On the other hand, clade 171K viruses have done remarkably well throughout 2016 and now comprise an estimated 69\% of currently circulating H3N2 viruses. This steady and rapid increase is strongly suggestive of an adaptive origin. Notably, 171K emerged and spread first in Asia, reaching nearly 80\% frequency in April 2016. Higher frequency of 171K in Asia is expected to spread to the rest of the world given historic geographic observations \cite{bedford2015global}. Within 171K, the variant 121K has spread. However, it doesn't appear to be spreading more rapidly than its parent 171K clade. Phylogenetic patterns suggest 171K as driver rather than 121K. Still, the rate of increase of 121K suggests that a sizable fraction of 2016-2017 viruses will be comprised of 121K.

@@

\textit{Without strong competition from another novel H3N2 virus, we believe that 171K will continue to increase in frequency in the global population and predominate in the 2016-2017 influenza season.} The continued spread of 171K is fully in line with the predictions we made in Feb 2016 \cite{feb2016report}. At this point, we can't say whether the 171K mutation in HA1 or the HA2 mutations 77V/155E (or some combination) is driving the selective spread of this clade.

@@

Analysis of local branching index \cite{neher2014predicting} also flags the clade 6b.1 as rapidly expanding and the most successful lineages within H1N1pdm (Fig.\ \ref{H1N1pdm_LBI}).

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/sep-2016/H1N1pdm_LBI.png}
	\caption{\textbf{H1N1pdm phylogeny colored by local branching index.}
	}
	\label{H1N1pdm_LBI}
\end{figure}

Analysis of antigenic evolution via the method of Neher et al. (2015) \cite{neher2015prediction} and using recent HI data from the Feb 2016 VCM report \cite{crickfeb2016report} from the Crick Institute Collaborating Center suggests a small effect on HI titer in both the 6b.1 and 6b.2 clades (Fig.\ \ref{H1N1pdm_HI}). In this case, 6b.1 shows an antigenic impact of 0.65 log2 units and 6b.2 shows an impact of 0.66 log2 units. Interesting, this analysis places the impact on clade 6b.1 from the 84N mutation rather than the later 162N or 216T mutations.

\begin{figure}[h!]
	\centering
	\includegraphics[width=1.0\textwidth]{../figures/sep-2016/H1N1pdm_HI.png}
	\caption{\textbf{H1N1pdm phylogeny colored by antigenicity.}
	The left panel shows individual HI measurements of viruses against serum from A/HongKong/5659/2012.
	Cooler color indicates great antigenic similarity (less titer drop going from homologous to heterologous titers).
	The right panel shows model estimates of antigenic divergence from A/HongKong/5659/2012.
	These estimates combine HI measurements from many different sera.
	Data is current through Feb 2016.
	}
	\label{H1N1pdm_HI}
\end{figure}

@@

\textbf{Clade 3 has continued to predominate the B/Yamagata population. Within clade 3 the 172Q/251V subclade has risen to predominate. There is limited diversity within 172Q/251V viruses, with the largest subclade comprised of 211R viruses.}
